---
title: "MobQoL OPUF value set GP vs MI"
date: "22 November 2022"
output: 
  md_document:
    toc: no
knit: (function(inputFile, encoding) { 
          rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file="README.md") })
---


This document contains the data, source code and main results for **"Developing preference-based value sets for the MobQoL-7D: Practical application of the Online Elicitation of Personal Utility Functions (OPUF) tool"** by Nathan Bray, Rhiannon Tudor Edwards, Paul Schneider.


<br>hr><br>

# Setup

```{r}
rm(list=ls())

set.seed(2022)

library(jsonlite)
library(ggplot2)
library(presize)
library(reshape2)

library(knitr)

# load utility functions
source("./R/mobqol_util.R")

# define dimension ID
ds_dims = c("AC","CO","PD","IN","SE","ME","AX")

# load eligible id 
eligible_ids_gp = read.csv("./data/gp_anonymised.csv")$id
eligible_ids_mi = read.csv("./data/mi_anonymised.csv")$id
# load raw response data
mq_jsons = read_json("./data/respones_anonymised.json")
# retrieve and process data from JSON
# N.B.: eligible ids are provided to exclude respondents who filled in
# both surveys (i.e. only including them to the one they filled in first)
gp = retrieveMobQoLPUFs(raw_jsons = mq_jsons, eligible_ids = eligible_ids_gp)
mi = retrieveMobQoLPUFs(mq_jsons, eligible_ids = eligible_ids_mi)
```


# Descriptive stats

Number of participants included in the analysis: **`r gp$total_n`** in the GP group; and **`r mi$total_n`** in the MI group.

 
# demographic characteristics

```{r results='asis'}
# Compare demographic characteristics between GP and MI group
group_vars = c("gender","age","any_mobility_problems","mobility_problems_meq_6m")
res_dual = c()
for(group in group_vars){
  # GP
  t_gp = table(gp$groups_df$value[gp$groups_df$group == group])
  tn_gp = names(t_gp)
  res_gp = paste(t_gp, " (", round(t_gp/length(unique(gp$groups_df$id[gp$groups_df$group == group])),3)*100,"%)", sep ="")
  names(res_gp) = tn_gp
  df_gp = data.frame(var = names(res_gp),value = res_gp)
  
  # MI
  t_mi = table(mi$groups_df$value[mi$groups_df$group == group])
  tn_mi = names(t_mi)
  res_mi = paste(t_mi, " (", round(t_mi/length(unique(mi$groups_df$id[mi$groups_df$group == group])),3)*100,"%)", sep ="")
  names(res_mi) = tn_mi
  df_mi = data.frame(var = names(res_mi),value = res_mi)
  
  df_dual = merge(df_gp, df_mi, by = "var", all = T,suffixes = c("GP","MI"))
  
  row_insert = cbind(group,"","")
  colnames(row_insert) = names(df_dual)
  
  res_dual_group = rbind(row_insert,df_dual)
  res_dual = rbind(res_dual, res_dual_group)
}
res_dual[is.na(res_dual)] = ""
print(kable(res_dual,col.names = c("Group","GP","MI")))
```



# Survey completion times

Mean completion time in the GP group = **`r round(mean(gp$time_df$seconds/60),1)` minutes (SD = `r round(sd(gp$time_df$seconds/60,1))`)**

Mean completion time in the MI group = **`r round(mean(mi$time_df$seconds/60),1)` minutes (SD = `r round(sd(mi$time_df$seconds/60,1))`)**


```{r}
kable(
  cbind(
    cbind(round(summary(round(gp$time_df$seconds/60,1)) ,1)),
    cbind(round(summary(round(mi$time_df$seconds/60,1)) ,1))
  ),
  col.names = c("GP", "MI"),
  caption = "Completion time (mins)"
  )
```



# Own health state

```{r}
# GP
fh_str = paste0(sum(gp$own_state_df$state == "1111111"), " (",round(sum(gp$own_state_df$state == "1111111")/gp$n_pufs*100),"%)")
only_minor = grepl("2",gp$own_state_df$state) &!grepl("3",gp$own_state_df$state) & !grepl("4",gp$own_state_df$state) 
only_minor_str = paste0(
  sum(only_minor), " (",
  round(sum(only_minor)/gp$n_pufs*100),"%)"
  )
severe = grepl("4",gp$own_state_df$state) 
severe_str = paste0(
  sum(severe), " (",
  round(sum(severe)/gp$n_pufs*100),"%)"
  )
severity_scores = unlist(lapply(strsplit(gp$own_state_df$state,""),\(x)sum(as.numeric(x))))
severity_scores_str = paste0(round(mean(severity_scores),1), " (", round(sd(severity_scores),1),")")

gp_hs = c(fh_str,only_minor_str,severe_str,severity_scores_str)

# MI
fh_str = paste0(sum(mi$own_state_df$state == "1111111"), " (",round(sum(mi$own_state_df$state == "1111111")/mi$n_pufs*100),"%)")
only_minor = grepl("2",mi$own_state_df$state) &!grepl("3",mi$own_state_df$state) & !grepl("4",mi$own_state_df$state)
only_minor_str = paste0(
  sum(only_minor), " (",
  round(sum(only_minor)/mi$n_pufs*100),"%)"
  )
severe = grepl("4",mi$own_state_df$state)
severe_str = paste0(
  sum(severe), " (",
  round(sum(severe)/mi$n_pufs*100),"%)"
  )
severity_scores = unlist(lapply(strsplit(mi$own_state_df$state,""),\(x)sum(as.numeric(x))))
severity_scores_str = paste0(round(mean(severity_scores),1), " (", round(sd(severity_scores),1),")")
  



mi_hs = c(fh_str,only_minor_str,severe_str,severity_scores_str)

hs_dual = cbind(gp_hs,mi_hs)
rownames(hs_dual) = c("No problems","Only minor","Severe in at least 1","Mean (SD) severity score")
colnames(hs_dual) = c("GP","MI")

kable(hs_dual,caption = "Reported health state profiles")

```

N.B. Minimum severity score = 7


# OPUF responses

# Dimension weights

```{r}
# GP
dim_weights_gp = by(gp$weights_df$value, gp$weights_df$dim_id, \(dim){
      paste(round(mean(dim),1), " (",round(sd(dim),1),")",sep = "")
})
dim_ranks_gp = rank(-by(gp$weights_df$value, gp$weights_df$dim_id, mean))
# MI
dim_weights_mi = by(mi$weights_df$value, mi$weights_df$dim_id, \(dim){
    paste(round(mean(dim),1), " (",round(sd(dim),1),")",sep = "")
})
dim_ranks_mi = rank(-by(mi$weights_df$value, mi$weights_df$dim_id, mean))
# combined
dim_weights_tbl = cbind(dim_weights_gp,dim_ranks_gp,dim_weights_mi,dim_ranks_mi)
rownames(dim_weights_tbl) = names(dim_weights_gp)
kable(dim_weights_tbl, col.names = c("GP Mean (SD)","GP Rank","MI Mean (SD)","MI Rank"))
```




```{r}
dim_weights_gp = by(gp$weights_df$value, gp$weights_df$dim_id, mean)
dim_weights_mi = by(mi$weights_df$value, mi$weights_df$dim_id, mean)

weights_df = data.frame(
  group = c(rep("GP",7),rep("MI",7)),
  dim = names(dim_weights_gp),
  value = c(dim_weights_gp,dim_weights_mi)
)

ggplot(weights_df, aes(fill=group, y=value, x=dim)) + 
    geom_bar(position="dodge", stat="identity", alpha = 0.7, color = "lightgray") +
  theme_minimal() +
  scale_fill_viridis_d(name = "Sample") +
  # theme(legend.position = "none") +
  ggtitle("Mean dimension weights by group")
```



# Level ratings

```{r}
lvl_ratings_gp = lapply(ds_dims,\(dim){
    s_ = grepl(dim,gp$ratings_df$lvl_id) & !grepl("_3",gp$ratings_df$lvl_id) & !grepl("_0",gp$ratings_df$lvl_id)
    df_ = gp$ratings_df[s_,]
    x = by(df_$value,df_$lvl_id,\(lvl){
        paste(round(mean(lvl),1), " (",round(sd(lvl),1),")",sep = "")
    })
    x = rbind(c(""),matrix(x))
    rownames(x) = c(dim,"lvl2","lvl3")
    x
  })
lvl_ratings_mi = lapply(ds_dims,\(dim){
    s_ = grepl(dim,mi$ratings_df$lvl_id) & !grepl("_3",mi$ratings_df$lvl_id) & !grepl("_0",mi$ratings_df$lvl_id)
    df_ = mi$ratings_df[s_,]
    x = by(df_$value,df_$lvl_id,\(lvl){
        paste(round(mean(lvl),1), " (",round(sd(lvl),1),")",sep = "")
    })
    x = rbind(c(""),matrix(x))
    rownames(x) = c(dim,"lvl2","lvl3")
    x
  })

lvl_ratings_tbl_gp = do.call(rbind, lvl_ratings_gp)
lvl_ratings_tbl_mi = do.call(rbind, lvl_ratings_mi)
lvl_ratings_tbl_dual = cbind(lvl_ratings_tbl_gp, lvl_ratings_tbl_mi)
kable(lvl_ratings_tbl_dual, col.names = c("GP Mean (SD)","MI Mean (SD)"))
```



# Anchoring

In the **GP** group, **`r sum((1-gp$anchor_df$anchor)>0)` (`r round(sum((1-gp$anchor_df$anchor)>0)/gp$n_pufs*100)`%)** Participants indicated they would prefer the worst state over being dead. Correspondingly, **`r sum((1-gp$anchor_df$anchor)<=0)` (`r round(sum((1-gp$anchor_df$anchor)<=0)/gp$n_pufs*100)`%)** Participants indicated they would prefer being dead over the worst state or be indifferent.

In the **MI** group, **`r sum((1-mi$anchor_df$anchor)>0)` (`r round(sum((1-mi$anchor_df$anchor)>0)/mi$n_pufs*100)`%)** Participants indicated they would prefer the worst state over being dead. Correspondingly, **`r sum((1-mi$anchor_df$anchor)<=0)` (`r round(sum((1-mi$anchor_df$anchor)<=0)/mi$n_pufs*100)`%)** Participants indicated they would prefer being dead over the worst state or be indifferent. 

```{r}
# GP 
# 'extreme preferences':
# sum(gp$anchor_df$anchor> 2) # n participants
# round(sum(gp$anchor_df$anchor> 2)/gp$n_pufs*100) # %
# Thus: trim anchor points to -1 (this is commonly done)
gp_trimmed = gp$anchor_df$anchor
gp_trimmed[gp_trimmed> 2] = 2
# with and without the outliers, new mean anchor point is different:
# 1-mean(gp_trimmed);sd(gp_trimmed);

# MI
# 'extreme preferences':
# sum(mi$anchor_df$anchor> 2) # n participants
# round(sum(mi$anchor_df$anchor> 2)/mi$n_pufs*100) # %
mi_trimmed = mi$anchor_df$anchor
mi_trimmed[mi_trimmed> 2] = 2

# distribution of achor points
ggplot() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_density(aes(1-gp_trimmed, fill = "GP"), col = "black", alpha = 0.5) +
  geom_density(aes(1-mi_trimmed, fill = "MI"), col = "black", alpha = 0.4) +
  scale_fill_viridis_d(name  = "Sample") +
  xlab("Density - distribution of utility values for the worst state") +
  theme_minimal()

```

After capping utlity scale at -1 (which is commonly done), the mean (SD) anchor point (i.e. the utility of the worst health state) was **`r 1-round(mean(gp_trimmed),3)` (`r round(sd(gp_trimmed),3)`)** in the GP group, and **`r round(1-mean(mi_trimmed),3)` (`r round(sd(mi_trimmed),3)`)** in the MI group.

N.B. In the GP group, `r sum(gp$anchor_df$anchor> 2)` (`r round(sum(gp$anchor_df$anchor> 2)/gp$n_pufs*100)` %) participants had anchor points below -1. In the MI group, there were `r sum(mi$anchor_df$anchor> 2)` (`r round(sum(mi$anchor_df$anchor> 2)/mi$n_pufs*100)` %) participants who had such low anchor points.



# MobQol value sets

```{r}
# Social preference model
# GP
gp_mean_est = formatC(apply(gp$pufs_trimmed,2,mean),digits = 3, format = "f")
gp_mean_boot = replicate(10000,apply(gp$pufs_trimmed[sample(1:nrow(gp$pufs_trimmed),size = nrow(gp$pufs_trimmed),replace = T),],2,mean))
gp_mean_boot = formatC(apply(gp_mean_boot,1,quantile,probs = c(0.025,0.975)),digits = 3, format = "f")
gp_m_95 = sapply(1:ncol(gp_mean_boot), \(i) paste(gp_mean_est[i]," (",gp_mean_boot[1,i],"; ",gp_mean_boot[2,i],")",sep = ""))

# MI
mi_mean_est = formatC(apply(mi$pufs_trimmed,2,mean),digits = 3, format = "f")
mi_mean_boot = replicate(10000,apply(mi$pufs_trimmed[sample(1:nrow(mi$pufs_trimmed),size = nrow(mi$pufs_trimmed),replace = T),],2,mean))
mi_mean_boot = formatC(apply(mi_mean_boot,1,quantile,probs = c(0.025,0.975)),digits = 3, format = "f")
mi_m_95 = sapply(1:ncol(mi_mean_boot), \(i) paste(mi_mean_est[i]," (",mi_mean_boot[1,i],"; ",mi_mean_boot[2,i],")",sep = ""))

# Dual
dual_m_95 = cbind(gp_m_95,mi_m_95)
rownames(dual_m_95) = paste0(rep(ds_dims, each = 4),"_",1:4)
kable(
  dual_m_95, 
  col.names = c("GP","MI"),
  caption = "Social (GP) and MI-based value set for the MobQoL - mean and 95%CI"
  )
```

# Value sets separately

## GP

```{r}
kable(matrix(
  gp_m_95,
  ncol = 7,
  dimnames = list(paste0("lvl_", 1:4), ds_dims)
),
caption = "GP",
align = "c")
```

## MI

```{r}
kable(matrix(
  mi_m_95,
  ncol = 7,
  dimnames = list(paste0("lvl_", 1:4), ds_dims)
),
caption = "MI",
align = "c")
```


# Value set difference

```{r}
# mean difference
m_diff = apply(gp$pufs_trimmed,2,mean) - apply(mi$pufs_trimmed,2,mean)
m_diff_str = formatC(m_diff,digits = 3, format = "f")

# bootstrapped 95% conf ints
boots = 10000
s_gp = replicate(boots, sample(1:nrow(gp$pufs_trimmed),replace = T))
s_mi = replicate(boots, sample(1:nrow(mi$pufs_trimmed),replace = T))
res = matrix(data = NA,ncol = ncol(mi$pufs_trimmed), nrow=boots )
for(i in 1:boots){
  s_gp_i = s_gp[,i]
  m_gp = apply(gp$pufs_trimmed[s_gp_i,],2,mean)
  
  s_mi_i = s_mi[,i]
  m_mi = apply(mi$pufs_trimmed[s_mi_i,],2,mean)
  
  res[i,] = m_gp - m_mi
  
}

res_str = apply(res,2,\(x){
  ci95 = quantile(x,c(0.025,0.975))
  ci95_str = formatC(ci95,digits = 3, format = "f")
  return(ci95_str)
})

res_str_mat = paste0(m_diff_str, " (",res_str[1,],"; ",res_str[2,],")")

kable(
  matrix(
    data = res_str_mat, 
    ncol = 7, 
    dimnames = list(paste0("lvl_",1:4),ds_dims)
    ), 
  caption = "GP - MI difference",
  align = "c"
  )
```




```{r}
# social value set
social_model = matrix(apply(gp$pufs_trimmed,2,mean), ncol = length(ds_dims))
ds_expanded = expand.grid(lapply(1:7, \(x) 1:4))
ds_str = apply(ds_expanded,1,\(x) paste0(x,collapse=""))
gp_vs = c()
for(i in 1:nrow(ds_expanded)){
  v_ = 1
  for(j in seq_along(ds_expanded[i,])){
    v_ = v_ - social_model[ds_expanded[i,j],j]
  }
  gp_vs = c(gp_vs, v_)
}

gp_hs_ranking = rank(-gp_vs)


mi_model = matrix(apply(mi$pufs_trimmed,2,mean), ncol = length(ds_dims))
mi_vs = c()
for(i in 1:nrow(ds_expanded)){
  v_ = 1
  for(j in seq_along(ds_expanded[i,])){
    v_ = v_ - mi_model[ds_expanded[i,j],j]
  }
  mi_vs = c(mi_vs, v_)
}

mi_hs_ranking = rank(-mi_vs)

dual_vs = data.frame(
  state = ds_str,
  gp = gp_vs,
  gp_vs_rank = gp_hs_ranking,
  mi = mi_vs,
  mi_rank = mi_hs_ranking
)
dual_vs$util_diff = dual_vs$gp - dual_vs$mi
dual_vs$rank_diff = dual_vs$gp_vs_rank - dual_vs$mi_rank
write.csv(dual_vs, file = "./output/dual_vs.csv")
# (see sample health state values below and full results in separate file 'output/dual_vs.csv')

```


# Number of states worse than dead?

Neither in the GP nor in the MI group is the utility value of any states negative. On average, both groups consider all health state better than dead.


# Value set visualisation
```{r}
ggplot() +
  geom_hline(yintercept = 0, col = "gray", size = 0.5) +
  # geom_line(aes(x = 1:length(gp_hs_ranking), y = gp_vs[gp_hs_ranking])) +
  geom_line(data = dual_vs, aes(x=gp_vs_rank, y = gp, col = "GP"), size = 0.8, alpha =.7) +
  geom_line(data = dual_vs, aes(x=gp_vs_rank, y = mi, col = "MI"), size = 0.5, alpha =.7) +
  scale_color_viridis_d(name = "Sample") +
  ylab("HRQoL / Utility") +
  xlab("MobQoL states - ranked from best to worst") +
  theme_minimal()
```


# Distribution of utility values
```{r}
  ggplot() +
  geom_histogram(data = dual_vs, aes(gp, fill = "GP"),color="darkgray", size = 0.8, alpha =.9, binwidth = 0.01) +
  geom_histogram(data = dual_vs, aes(mi, fill = "MI"),color="gray", size = 0.8, alpha =.6,binwidth = 0.01) +
  scale_fill_viridis_d(name = "Sample") +
  ylab("Frequency") +
  xlab("HRQoL / Utility") +
  theme_minimal()
```


# Value set table - GP vs MI, top and bottom 100 states (of all 16,384 states)
```{r}
dual_vs = dual_vs[order(dual_vs$gp_vs_rank),]
kable(
  rbind(head(dual_vs, 100),tail(dual_vs, 100)), 
  col.names = c("Health state","GP utility","GP rank","MI utility", "MI rank","Util diff","Rank diff"),
  caption = "GP vs MI value sets",
  row.names = F
  )
```



